---
title: "Project 1, Part 2 of Statistical Inference"
author: "Jin-Keat Lim"
date: "Friday, October 24, 2014"
output: pdf_document
---
**Introduction**

Before beginning the project, I simply loaded the ToothGrowth dataset into a dataframe. Note I hid the warnings that both "ggplot2" and "xtable" packages were built under R version 3.1.1.
```{r Proj1Pt1, results="hide", warning=FALSE}
library(ggplot2)
library(xtable)
df <- data.frame(ToothGrowth)
```

**Part 1** of the project - basic Exploratory Data Analysis of the dataset
```{r, fig.show="hold", fig.height=2, fig.width=2, dev.args=list(pointsize=7)}
hist(df$dose)
hist(df$len)

print(qplot(dose, len, data=df, color=supp, fill=supp) + theme(legend.position="bottom"))
print(qplot(dose, len, data=df, facets = .~supp))
print(qplot(supp, len, data=df, facets = .~dose))
```

Using various histograms and breakdowns, it is easy to see that the doses are separated in 3 categories. Further analysis will be needed to evaluate the effect of **supp** and **dose** on the ToothGrowth **len**.

\newpage

**Part 2** of the project - a basic summary of the data
```{r, results="asis"}
options(xtable.comment = FALSE)
print(xtable(summary(df), caption="Summary of entire dataset"))
print(xtable(summary(df[df$supp=="OJ",]), caption="Summary of OJ supp"))
print(xtable(summary(df[df$supp=="VC",]), caption="Summary of VC supp"))

```
**Part 3** of the project - using confidence intervals and hypothesis tests to compare tooth growth by supp and dose. 
```{r, results="asis"}
VC_df <- df[df$supp=="VC",]
OJ_df <- df[df$supp=="OJ",]

#note I am using T-CI and the qt() function rather than using qnorm()

##T-CI by supp
OJ_len_CI <- mean(OJ_df$len) + c(-1,1) * qt(0.975, length(OJ_df$len)-1) + 
              sd(OJ_df$len) / sqrt(length(OJ_df$len))
VC_len_CI <- mean(VC_df$len) + c(-1,1) * qt(0.975, length(VC_df$len)-1) + 
              sd(VC_df$len) / sqrt(length(VC_df$len))

CI_df <- t(data.frame(OJ_len_CI, VC_len_CI))
colnames(CI_df) <- c("lower limit", "upper limit")
xtable(CI_df, caption="Summary of T-CI lengths by supp")

```

```{r, results="asis"}
##T-CI by dose
dose1_df <- df[df$dose==0.5,];dose2_df <- df[df$dose==1,];dose3_df <- df[df$dose==2,]

dose1_len_CI <- mean(dose1_df$len) + c(-1,1) * qt(0.975, length(dose1_df)-1) + 
  sd(dose1_df$len) / sqrt(length(dose1_df$len))
dose2_len_CI <- mean(dose2_df$len) + c(-1,1) * qt(0.975, length(dose2_df)-1) + 
  sd(dose2_df$len) / sqrt(length(dose2_df$len))
dose3_len_CI <- mean(dose3_df$len) + c(-1,1) * qt(0.975, length(dose3_df)-1) + 
  sd(dose3_df$len) / sqrt(length(dose3_df$len))

CI_dose_df <- t(data.frame(dose1_len_CI, dose2_len_CI, dose3_len_CI))
colnames(CI_dose_df) <- c("lower limit", "upper limit")
xtable(CI_dose_df, caption="Summary of T-CI lengths by dose")

```
For easier visualization and to also help determine a relevant hypothesis test, I plotted the density of the ToothGrowth **len** broken down by **supp** and **dose**:
```{r, fig.show="hold", fig.height=3, fig.width=3, dev.args=list(pointsize=8)}
#taking a look at the density of the tooth growth broken down by dose
plot_dose <- ggplot(df, aes(x=df$len, fill=factor(df$dose))) +
        geom_density(alpha=0.2, binwidth=1) + 
        theme(legend.position="bottom")
print(plot_dose)

#taking a look at the density of the tooth growth broken down by supp
plot_supp <- ggplot(df, aes(x=df$len, fill=factor(df$supp))) + 
  geom_density(alpha=0.2, binwidth=1) + 
  theme(legend.position="bottom")
print(plot_supp)
```

Based on these plots, we can subjectively see that the **dose** affects the **len**, and the effect of the **supp** on the **len** is less clear. I decided to first construct a hypothesis test to test the effect of **supp** on **len**: 

**Null hypothesis:**the mean length of group **supp** OJ is equal to the mean length of group **supp** VC
```{r}
print(t.test(len ~ supp, paired=FALSE, var.equal=FALSE, data=df))
```

The p-value is 6.06%, and is the probability of the null being true. Therefore, this can be considered a failure to reject the null hypothesis (assuming alpha is 5% using a standard 95% CI). 

Because there are three discrete **dose** values, constructing a hypothesis test to test the effect of **dose** on **len** require testing of the below null hypotheses:

- *H*~A~: mean length of group **dose** 0.5 is equal to the mean length of group **dose** 1.0
- *H*~B~: mean length of group **dose** 1.0 is equal to the mean length of group **dose** 2.0
- *H*~C~: mean length of group **dose** 0.5 is equal to the mean length of group **dose** 2.0

```{r, results="hold"}
#testing Ha, Hb, Hc, respectively
t.test(len ~ dose, paired=FALSE, var.equal=FALSE, data= subset(df, dose %in% c(0.5,1)))
t.test(len ~ dose, paired=FALSE, var.equal=FALSE, data= subset(df, dose %in% c(1,2)))
t.test(len ~ dose, paired=FALSE, var.equal=FALSE, data= subset(df, dose %in% c(0.5,2)))
```

All 3 of these tests show small p-values, and the null hypotheses *H*~A~, *H*~B~, *H*~C~ can all be rejected. We can conclude the alternative hypotheses that the **dose** administered does make a difference in the **len**.  


**Part 4 - Conclusion and Assumptions**

For all hypothesis testing, an alpha of 0.05 was assumed (standard 95% CI). In addition, each **t.test()** assumed unequal variances between each corresponding group. 

When testing the effect of **supp** on **len**, the p-value 0.0606 > alpha 0.05, resulting in a failure to reject the null hypothesis. There is not sufficient evidence to say that the **supp** administered makes no difference in the **len**. 

When testing the effect of **dose** on **len**, the *H*~A~, *H*~B~, *H*~C~ p-values  << alpha 0.05, resulting in a rejection of the null hypothesis. We can state there is sufficient evidence to reject the null hypotheses, and accept the alternate hypotheses that the **dose** administered makes a difference in the **len**. This is supported by the density plot observed earlier.